Type Safe Redis Queries

### 2017-07-13 TyDe2017/sections/IndexedMonads-orig.tex
Auto stash before merge of "master" and "origin/master"

Stateful computations are often reasoned using Hoare logic. A {\em Hoare triple}
$\{P\} S \{Q\}$ denotes such a proposition: if the statement $S$ is executed in
a state satisfying prediate $P$, when it terminates, the state must satisfy
predicate $Q$. Predicates $P$ and $Q$ are respectively is called the
\emph{precondition} and the \emph{postcondition} of the Hoare triple.

In Haskell, stateful computations are represented by monads. In order to
reason about their behaviors within the type system, we wish to label a state
monad with its pre and postcondition. An \emph{indexed monad}~%
\cite{indexedmonad} (also called \emph{monadish} or \emph{parameterised monad})
is a monad that, in addition to the type of value it computes, takes two more
type arguments representing an initial state and a final state, to be
interpreted like a Hoare triple~\cite{kleisli}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The intention is that a computation of type \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is a stateful computation
such that, if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}. The operator \ensuremath{\Varid{unit}}
lifts a pure computation to a stateful computation that does not alter the
state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is followed by
\ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}} --- the postcondition of \ensuremath{\Varid{x}} matches the precondition of
the computation returned by \ensuremath{\Varid{f}}. The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}}.
Indexed monads have been used ~\cite{typefun,staticresources} ... \todo{for what? Some discriptions here to properly cite them.}

We define a new indexed monad \ensuremath{\Conid{Edis}} which, at term level, merely wraps
\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the
pre/postconditions at type level:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Edis}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Edis}\;\{\mskip1.5mu \Varid{unEdis}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Edis}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Edis}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Edis}\;(\Varid{unEdis}\;\Varid{m}\bind \Varid{unEdis}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}} methods are not interesting: they merely
make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Edis}} when necessary. With \ensuremath{\Conid{Edis}} being a \ensuremath{\mathbf{newtype}}, they
can be optimized away in runtime. The interesting bits happen in compile type,
on the added type information.

The properties of the state we care about are the set of currently allocated
keys and their associated types. We will present, in Section~\ref{sec:type-level-dict}, techniques that allow us to specify
properties such as ``the keys in the database are \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}},
respectively associated to values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.''
For now, however, let us look at the simplest \Redis{} command.

The command \text{PING} in \Redis{} does nothing but replies a message
\text{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Edis{} version of \ensuremath{\Varid{ping}} simply
applys an additional constructor (functions from \Hedis{} are qualified with
\ensuremath{\Conid{Hedis}} to prevent name clashing):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Edis}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Since \ensuremath{\Varid{ping}} does not alter the database, the postcondition and precondition
are the same. Commands that are more interesting will be introduced after
we present our type-level encoding of constraints on states.

>>>>

Stateful computations are often reasoned using Hoare logic. A {\em Hoare triple}
$\{P\} S \{Q\}$ denotes the following proposition: if the statement $S$ is
executed in a state satisfying predicate $P$, when it terminates, the state must
satisfy predicate $Q$. Predicates $P$ and $Q$ are respectively called the
\emph{precondition} and the \emph{postcondition} of the Hoare triple.

In Haskell, stateful computations are represented by monads. To reason about
them within the type system, we wish to label a state monad with its pre and
postcondition. An \emph{indexed monad}~\cite{indexedmonad} (also called a
\emph{parameterised monad} or \emph{monadish}) is a monad that, in addition to
the type of its result, takes two more type parameters representing an initial
state and a final state, to be interpreted like a Hoare triple:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The intention is that a computation of type \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is a stateful computation
such that, if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}. The operator \ensuremath{\Varid{unit}}
lifts a pure computation to a stateful computation that does not alter the
state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is followed by
\ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}} --- the postcondition of \ensuremath{\Varid{x}} matches the precondition of
the computation returned by \ensuremath{\Varid{f}}. The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}}.

We define a new indexed monad \ensuremath{\Conid{Edis}}. At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}}
methods are not interesting: they merely make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of
\ensuremath{\Conid{Redis}}, and extract and re-apply the constructor \ensuremath{\Conid{Edis}} when necessary.
With \ensuremath{\Conid{Edis}} being a \ensuremath{\mathbf{newtype}}, they can be optimized away in runtime. The
purpose is to add the pre/postconditions at type level:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Edis}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Edis}\;\{\mskip1.5mu \Varid{unEdis}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Edis}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Edis}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Edis}\;(\Varid{unEdis}\;\Varid{m}\bind \Varid{unEdis}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The properties of the state we care about are the set of currently
allocated keys and types of their values. We will present, in Section~\ref{sec:type-level-dict}, techniques that allow us to specify
properties such as ``the keys in the data store are \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}},
respectively assigned values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.'' For now,
however, let us look at the simplest \Redis{} command.

\Redis{} commands can be executed in two contexts: the normal context, or a \emph{transaction}. In \Hedis{}, a command yielding value of type \ensuremath{\Varid{a}} in the
normal case is represented by \ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Varid{a})}, as mentioned in
Section~\ref{sec:introduction}; in a transaction, the command is represented by
two other datatypes \ensuremath{\Conid{RedisTx}\;(\Conid{Queued}\;\Varid{a})}. In this paper we focus on the former
case. For brevity we abbreviate \ensuremath{\Conid{Either}\;\Conid{Reply}\;\Varid{a}} to \ensuremath{\Conid{R} \uplus \Varid{a}}.

The command \texttt{PING} in \Redis{} does nothing but replying a message
\texttt{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
\ensuremath{\Conid{Redis}\;(\Conid{R} \uplus \Conid{Status})}. The \Edis{} version of \ensuremath{\Varid{ping}} simply applies an additional constructor (functions from \Hedis{} are qualified with \ensuremath{\Conid{Hedis}} to
prevent name clashing):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Edis}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Since \ensuremath{\Varid{ping}} does not alter the data store, the postcondition and precondition
are the same. Commands that are more interesting will be introduced after
we present our type-level encoding of constraints on states.

### 2017-07-13 TyDe2017/sections/Introduction-orig.tex
Auto stash before merge of "master" and "origin/master"

\Redis{}\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as database, cache and message broker. A \Redis{} datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
\Redis{} allows values of various types, including strings, hashes, lists, and sets, etc, to be stored, and provides a collection of of atomic \emph{commands} to manipulate these values.

For an example, consider the following sequence of commands, entered through the interactive interface of \Redis{}. The keys \texttt{some-set} and \texttt{another-set}
are both associated to a set. The two call to command \texttt{SADD} respectively
adds three and two values to the two sets, before \texttt{SINTER} takes their intersection:
\begin{tabbing}\tt
~redis\char62{}~SADD~some\char45{}set~a~b~c\\
\tt ~\char40{}integer\char41{}~3\\
\tt ~redis\char62{}~SADD~another\char45{}set~a~b\\
\tt ~\char40{}integer\char41{}~2\\
\tt ~redis\char62{}~SINTER~some\char45{}set~another\char45{}set\\
\tt ~1\char41{}~\char34{}a\char34{}\\
\tt ~2\char41{}~\char34{}b\char34{}
\end{tabbing}

Note that the keys \texttt{some-set} and \texttt{another-set}, if not existing before the call to \texttt{SADD}, are created on site. The call to
\texttt{SADD} returns the size of the set after completion of the command.

Many third party libraries provide interfaces that allow general purpose programming languages to access \Redis{} through its TCP protocol.
For Haskell, the most popular library is \Hedis{}\footnote{\url{https://hackage.haskell.org/package/hedis}}.
The following program implements the previous example:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 another-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34},\text{\tt \char34 c\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The function \ensuremath{\Varid{sadd}\mathbin{::}\Conid{ByteString}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer})} takes a key and a list of values as arguments, and returns
an \ensuremath{\Conid{Integer}} on success, or returns a \ensuremath{\Conid{Reply}}, a low-level representation of
replies from the Redis server, in case of failures. All wrapped in the monad
\ensuremath{\Conid{Redis}}, the context of command execution.\footnotemark

Note that keys and values, being nothing but binary strings in Redis, are
represented using Haskell \ensuremath{\Conid{ByteString}}. Values of other types must be encoded
as \ensuremath{\Conid{ByteString}}s before being written to the database, and decoded after being
read back.

\footnotetext{\Hedis{} provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class \ensuremath{\Conid{RedisCtx}}. We
demonstrate only \ensuremath{\Conid{Redis}} in this paper.}

%\paragraph{Motivation}
%All binary strings are equal, but some binary strings are more equal than others.
%While everything in \Redis{} is essentially a binary string, these strings
%are treated differently.
\Redis{} supports many different kind of data
structures, such as strings, hashes, lists, etc. While they are all encoded as
binary strings before being written to the database, most commands only works
with data of certain types. In the following example, the key
\texttt{some-string} is associated to string \texttt{foo} --- the command
\texttt{SET} always associates a key to a string. The subsequent call to \texttt{SADD}, which adds a value to a set, thus causes a runtime error.
\begin{tabbing}\tt
~redis\char62{}~SET~some\char45{}string~foo\\
\tt ~OK\\
\tt ~redis\char62{}~SADD~some\char45{}string~bar\\
\tt ~\char40{}error\char41{}~WRONGTYPE~Operation~against~a~key~holding~the~wrong\\
\tt ~kind~of~value
\end{tabbing}

%\paragraph{Example 2} Even worse, not all strings are equal!
% The call \texttt{INCR some-string} parses the string associated with key
% \texttt{some-string} to an integer, increments it by one, and store it back as
% a string. If the string can not be parse as an integer, a runtime error
% is raised.
% \begin{verbatim}
% redis> SET some-string foo
% OK
% redis> INCR some-string
% (error) ERR value is not an integer or out of range
% \end{verbatim}

Being a simple wrapper on top of the TCP protocol of \Redis{}, \Hedis{}
inherits the problem. Executing following program yields the same error
wrapped in Haskell: \ensuremath{\Conid{Left}\;(\Conid{Error}} \texttt{"WRONGTYPE Operation against a
key holding the wrong kind of value"}\ensuremath{)}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{set}\;\text{\tt \char34 some-string\char34}\;\text{\tt \char34 foo\char34}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-string\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

% \paragraph{The Cause} Every key is associated with a value, and every value has
% its own type. But most commands in \Redis{} only work with a certain type of
%  value. When a command is used on a wrong type of key, a runtime error occurs.
%  The problems illustrated above arise from the absence of type checking, with
%  respects to \textbf{the type of a value that associates with a key}.
%  These problems could have been avoided, if we could know the type every key
%  associates with in advance, and prevent programs with invalid commands from
%  executing.
%
% \paragraph{Hedis as an embedded DSL}
% Haskell makes it easy to build and use {\em domain specific embedded languages} (DSELs), and \Hedis{} can be regarded as one of them. What makes \Hedis{} peculiar is that,
%  it has \emph{variable bindings} (between keys and values), but with very
%  little or no semantic checking, neither dynamically nor statically.
%
Such a programming model is certainly very error-prone. Working within Haskell,
a host language with a strong typing system, one naturally wishes to build a
a domain-specific embedded language (DSEL) that exploits the rich type system
of Haskell to ensure absence of \Redis{} type errors.

This paper discusses the techniques we used and experiences we learned from building such a language, nicknamed \Edis{}. We constructed an {\em indexed
monad}, on top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that
maintains the set of currently defined keys and their types. To represent
the dictionary, we need to encode variable binds with {\em type-level} lists
and strings. To summarize our contributions:
\begin{itemize}
\item We present \Edis{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}.
% also makes Redis polymorphic by automatically converting back and forth from values of arbitrary types and boring ByteStrings.
%
\item We demonstrate how to model variable bindings of an embedded DSL using
 language extensions including type-level literals and data kinds.
%
\item We provide (yet another) an example of encoding effects and constraints of
in types, with indexed monad~\cite{indexedmonad}, closed type-families~\cite{closedtypefamilies} and constraints kinds~\cite{constraintskinds}.
\end{itemize}
\todo{Phrase this better.}

>>>>

\Redis{}\footnote{\url{https://redis.io}} is an open source, in-memory data
structure store that can be used as database, cache, and message broker, etc.
A \Redis{} data store can be thought of as a set of key-value pairs. The value
can be a string, a list of strings, a set of strings, or a hash table of
strings, etc. However, string is the only primitive datatype. Numbers, for
example, are serialized to strings before being saved in the data store, and
parsed back to numbers to be manipulated with. While the concept is simple,
\Redis{} is used as an essential component in a number of popular, matured services, including Twitter, GitHub, Weibo, StackOverflow, and Flickr, etc.

For an example, consider the following sequence of commands, entered through the
interactive interface of \Redis{}. The keys \texttt{some-set} and
\texttt{another-set} are both assigned a set of strings. The two calls to
command \texttt{SADD} respectively add three and two strings to the two sets,
before \texttt{SINTER} takes their intersection:
\begin{Verbatim}[xleftmargin=.4in]
redis> SADD some-set a b c
(integer) 3
redis> SADD another-set a b
(integer) 2
redis> SINTER some-set another-set
1) "a"
2) "b"
\end{Verbatim}
\noindent Notice that the keys \texttt{some-set} and \texttt{another-set}, if
not existing before the call to \texttt{SADD}, are created on site. The calls to
\texttt{SADD} return the sizes of the resulting sets.

Many third party libraries provide interfaces for general purpose programming
languages to access \Redis{} through its TCP protocol. For Haskell, the most
popular library is
\Hedis{}.\footnote{\url{https://hackage.haskell.org/package/hedis}}
A (normal) \Redis{} computation returning a value of type \ensuremath{\Varid{a}} is represented in
Haskell by \ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Varid{a})}, where the type \ensuremath{\Conid{Redis}} is a monad, while
\ensuremath{\Conid{Either}\;\Conid{Reply}\;\Varid{a}} indicates that the computation either returns a value of type
\ensuremath{\Varid{a}}, or fails with an error message represented by type \ensuremath{\Conid{Reply}}. The following
program implements the previous example:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 another-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34},\text{\tt \char34 c\char34}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The function \ensuremath{\Varid{sadd}\mathbin{::}\Conid{ByteString}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer})} takes a key and a list of values as arguments, and returns a
\Redis{} computation yielding \ensuremath{\Conid{Integer}}. Keys and values, being nothing but
binary strings in \Redis{}, are represented using Haskell \ensuremath{\Conid{ByteString}}.

\paragraph{The Problems} Most commands work only with data of certain types. In
the following example, the key \texttt{some-string} is assigned a string
\texttt{foo} --- the command \texttt{SET} always assigns a string to a key.
The subsequent call to \texttt{SADD}, which adds a value to a set, thus raises a
runtime error.
\begin{Verbatim}[xleftmargin=.4in]
redis> SET some-string foo
OK
redis> SADD some-string bar
(error) WRONGTYPE Operation against a key holding
the wrong kind of value
\end{Verbatim}
\noindent For another source of type error, the command \texttt{INCR key}
increments the value associated with \texttt{key} by one. With strings being the
only primitive type, \Redis{} parses the stored string to an integer and, after
incrementation, stores a string back. If the string cannot be parsed as an
integer, a runtime error is raised.

The reader must have noticed the peculiar pattern of value creation and updating
in \Redis{}: the same command is used both to create a key-value pair and to
update them. Similar to \texttt{SADD}, the command \texttt{LPUSH} appends a
value (a string) to a list, or creates one if the key does not exist:
\begin{Verbatim}[xleftmargin=.4in]
redis> LPUSH some-list bar
(integer) 1
\end{Verbatim}
\noindent Another command \texttt{LLEN} returns the length of the list, and
signals an error if the key is not associated with a list:
\begin{Verbatim}[xleftmargin=.4in]
redis> LLEN some-list
(integer) 1
redis> SET some-string foo
OK
redis> LLEN some-string
(error) WRONGTYPE Operation against a key holding
the wrong kind of value
\end{Verbatim}
\noindent Curiously, however, when applied to a key not yet created, \Redis{}
designers chose to let \texttt{LLEN} return \texttt{0}:
\begin{Verbatim}[xleftmargin=.4in]
redis> LLEN nonexistent
(integer) 0
\end{Verbatim}

Being a simple wrapper on top of the TCP protocol of \Redis{}, \Hedis{}
inherits all the behaviors. Executing the following program yields the same
error, but wrapped in a Haskell constructor: \ensuremath{\Conid{Left}\;(\Conid{Error}} \texttt{"WRONGTYPE
Operation against a key holding the wrong kind of value"}\ensuremath{)}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{set}\;\text{\tt \char34 some-string\char34}\;\text{\tt \char34 foo\char34}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-string\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Such a programming model is certainly very error-prone. Working within Haskell,
a host language with a strong typing system, one naturally wishes to build a
domain-specific embedded language (DSEL) that exploits the rich type system
of Haskell to not only ensure the absence of \Redis{} type errors, but also
provides better documentations. We wish to be sure that a program calling
\texttt{INCR}, for example, can be type checked only if we can statically
guarantee that the value to be accessed is indeed an integer. We wish to see
from the type of operators such as \texttt{LLEN} when it can be called, and
allow it to be used only in contexts that are safe. We may even want to
explicitly declare a fresh key and its type, to avoid reusing an existing key
by accident, and to prevent it from unexpectedly being used as some other type.

This paper discusses the techniques we used and the experiences we learned from
building such a language, \Edis{}. We constructed an {\em indexed monad}, on
top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that maintains the
set of currently defined keys and their types. To represent the dictionary, we
need to encode variable bindings with {\em type-level} lists and strings. And
to manipulate the dictionary, we applied various type-level programming
techniques. To summarize our contributions:
\begin{itemize}
\item We present \Edis{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}. Serializable Haskell datatypes are
automatically converted before being written to \Redis{} data store. Available
keys and their types are kept track of in type-level dictionaries. The types of
embedded commands state clearly their preconditions and postconditions on the
available keys and types, and a program is allowed to be constructed only if
it is guaranteed not to fail with a type error.
%
\item We demonstrate the use of various type-level programming techniques,
including data kinds, singleton types and proxies, closed type families, etc.,
to define type-level lists and operations that observe and manipulate the lists.
%
\item This is (yet another) example of encoding effects and constraints of programs in types, using indexed monad~\cite{indexedmonad}, closed
type-families~\cite{tfclosed} and constraint kinds~\cite{constraintskinds}.
\end{itemize}

In Section~\ref{sec:indexed-monads} we introduce indexed monads, to reason about
pre and postconditions of stateful programs, and in Section~\ref{sec:type-level-dict} we review the basics of type-level
programming in Haskell that allows us to build type-level dictionaries to keep
track of keys and types. Embeddings of \Redis{} commands are presented in
Section~\ref{sec:embedding-commands}. In Section~\ref{sec:discussions} we
discuss various issues regarding design choices, limitations of this approach,
as well as possible future works, before we review related work in
Section~\ref{sec:conclusions}.
